---
title: "Statistical Inference Course Project, Part I"
author: "Allen Hammock"
date: "October 12, 2014"
output: html_document
---
```{r compile_opts, echo = FALSE}
    SHOW.ALL <- FALSE
    suppressMessages(library(knitr))
    opts_knit$set(verbose = SHOW.ALL, progress = TRUE)
```

```{r initialize, echo = SHOW.ALL | FALSE}
# set the working directory
setwd("~/Google Drive/projects/r/coursera/statinference-006")

# pre-load all required dependencies
load.package <- function(x) {
    if (!suppressMessages(require(x, character.only = TRUE))) {
        warning(paste("WARNING. Package", x, "not found. Installing now."))
        install.packages(x, dep = TRUE)
        if(!require(x, character.only = TRUE)) {
            stop(paste("ABORT. Package", x, "not found"))
        }
    }
}

required.packages <- c("dplyr", "ggplot2")
for(i in 1:length(required.packages)) {
    load.package(required.packages[i])   
}
```
# Background

The exponential distribution can be simulated in R with `rexp(n, lambda)` where `lambda` is the rate parameter. The mean of exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also also $\frac{1}{\lambda}$. Set $\lambda={0.2}$ for all of the simulations. In this simulation, you will investigate the distribution of averages of 40 exponential(0.2)s. Note that you will need to do a thousand or so simulated averages of 40 exponentials.

## Part 1

### Problems
Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponential(0.2)s. 

1. Show where the distribution is centered at and compare it to the theoretical center of the distribution.

2. Show how variable it is and compare it to the theoretical variance of the distribution.

3. Show that the distribution is approximately normal.

4. Evaluate the coverage of the confidence interval for $\frac{1}{\lambda}$: $\overline{X}\pm{1.96}\frac{s}{\sqrt{n}}$.


```{r finalize, echo = SHOW.ALL | FALSE}
# save working image so we can look at it later
save.image("~/Google Drive/projects/r/coursera/statinference-006/statinference-006.RData")
```

### Evaluation notes

*Regarding the distribution of the mean of 40 exponentials. Did the student show where the distribution is centered at and compare it  to the theoretical center of the distribution?*

0: No<br />
1: Yes

*Regarding the distribution of the mean of 40 exponentials. Did the student show how variable it is and compare it to the theoretical variance of the distribution?*

0: No<br />
1: Yes

*Regarding the distribution of the mean of 40 exponentials. Did the student show that the distribution is approximately normal?*

0: No<br />
1: Yes

*Did the students evaluate  the coverage of the confidence interval for 1/lambda: $\overline{X}\pm{1.96}\frac{s}{\sqrt{n}}$*

0: No<br />
1: Yes

*Was the report created in knitr?*

0: No<br />
1: Yes

*Here's your opportunity to give this project +1 for effort. Did the student basically try to answer the question?*

0: No, it was really a cursory effort<br />
1: Yes, the student put in real effort and made an attempt t (regardless of correctness)

## Part 2

### Problems

Now in the second portion of the class, we're going to analyze the ToothGrowth data in the R datasets package. 

1. Load the ToothGrowth data and perform some basic exploratory data analyses 
2. Provide a basic summary of the data.
3. Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose. (Use the techniques from class even if there's other approaches worth considering)
4. State your conclusions and the assumptions needed for your conclusions. 

### Evalutation notes

*Did the student perform an exploratory data analysis?*

0: No<br />
1: Yes

*Did the student give a basic information summarizing the data set?*

0: No<br />
1: Yes

*Did the student perform some relevant confidence intervals and/or tests?*

0: No<br />
1: Yes

*Were the results of the tests and/or intervals interpreted in the context of the problem correctly? *

0: No<br />
1: Yes

*Did the student investigate the assumptions needed for their conclusions?*

0: No<br />
1: Yes

*Were multiple comparisons discussed appropriately?*

0: No<br />
1: Yes

*Here's your opportunity to give this project +1 for effort. Did the student basically try to answer the question?*

0: No it really was a poor effort<br />
1: Yes, this was a nice attempt (regardless of correctness)
