---
title: "Statistical Inference Course Project, Part 2"
author: "Allen Hammock"
date: "October 12, 2014"
output: html_document
---
```{r compile_opts, echo = FALSE}
    SHOW.ALL <- FALSE
    suppressMessages(library(knitr))
    opts_knit$set(verbose = SHOW.ALL, progress = TRUE)
```

```{r initialize, echo = SHOW.ALL | FALSE}
# set the working directory
setwd("~/Google Drive/projects/r/coursera/statinference-006")

# pre-load all required dependencies
load.package <- function(x) {
    if (!suppressMessages(require(x, character.only = TRUE))) {
        warning(paste("WARNING. Package", x, "not found. Installing now."))
        install.packages(x, dep = TRUE)
        if(!require(x, character.only = TRUE)) {
            stop(paste("ABORT. Package", x, "not found"))
        }
    }
}

required.packages <- c("dplyr", "ggplot2")
for(i in 1:length(required.packages)) {
    load.package(required.packages[i])   
}
```
# Background

## Part 2

### Problems

Now in the second portion of the class, we're going to analyze the ToothGrowth data in the R datasets package. 

1. Load the ToothGrowth data and perform some basic exploratory data analyses.

```{r pt2_1, echo = SHOW.ALL | TRUE, cache = FALSE}
# set the seed so others can reproduce easily
alicia.keys <- 4894608
set.seed(alicia.keys)

```

2. Provide a basic summary of the data.

```{r pt2_2, echo = SHOW.ALL | TRUE, cache = FALSE}
# set the seed so others can reproduce easily
mike.jones <- 2818004
set.seed(mike.jones)

```

3. Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose. (Use the techniques from class even if there's other approaches worth considering)

```{r pt2_3, echo = SHOW.ALL | TRUE, cache = FALSE}
# set the seed so others can reproduce easily
public.enemy <- 911
set.seed(public.enemy)

```

4. State your conclusions and the assumptions needed for your conclusions. 

```{r pt2_4, echo = SHOW.ALL | TRUE, cache = FALSE}
# set the seed so others can reproduce easily
soulja.boy <- 9998212
set.seed(soulja.boy)

```

### Evalutation notes

*Did the student perform an exploratory data analysis?*

0: No<br />
1: Yes

*Did the student give a basic information summarizing the data set?*

0: No<br />
1: Yes

*Did the student perform some relevant confidence intervals and/or tests?*

0: No<br />
1: Yes

*Were the results of the tests and/or intervals interpreted in the context of the problem correctly? *

0: No<br />
1: Yes

*Did the student investigate the assumptions needed for their conclusions?*

0: No<br />
1: Yes

*Were multiple comparisons discussed appropriately?*

0: No<br />
1: Yes

*Here's your opportunity to give this project +1 for effort. Did the student basically try to answer the question?*

0: No it really was a poor effort<br />
1: Yes, this was a nice attempt (regardless of correctness)


```{r finalize, echo = SHOW.ALL | FALSE}
# save working image so we can look at it later
save.image("~/Google Drive/projects/r/coursera/statinference-006/statinference-006-pt2.RData")
```