---
title: "Statistical Inference Course Project, Part I"
author: "Allen Hammock (brainvat)"
output: html_document
---
```{r compile_opts, echo = FALSE}
    SHOW.ALL <- FALSE
    suppressMessages(library(knitr))
    opts_knit$set(verbose = SHOW.ALL, progress = TRUE)
```

```{r initialize, echo = SHOW.ALL | FALSE}
# set the working directory
setwd("~/Google Drive/projects/r/coursera/statinference-006")

# pre-load all required dependencies
load.package <- function(x) {
    if (!suppressMessages(require(x, character.only = TRUE))) {
        warning(paste("WARNING. Package", x, "not found. Installing now."))
        install.packages(x, dep = TRUE)
        if(!require(x, character.only = TRUE)) {
            stop(paste("ABORT. Package", x, "not found"))
        }
    }
}

required.packages <- c("dplyr", "ggplot2")
for(i in 1:length(required.packages)) {
    load.package(required.packages[i])   
}
```

# The Exponential Distribution

## $$
    f(x)=\lambda e^{-\lambda x}
$$

The exponential distribution has its name because its probability distribution function has the shape of the exponential function. It crosses the Y-axis at some positive value called $\lambda$ and slopes downwards to the right, decreasing towards zero as the values of random variable $X$ increase, never reaching zero. The amount of slope in the curve is determined by the value of $\lambda$, also called *the rate*.

*Definition and graph adapted from http://www.ynegve.info/Post/169/a-tutorial-on-probability-and-exponential-distribution*

```{r fig0, echo = SHOW.ALL | TRUE, cache = FALSE}
x = seq(0, 40, length = 200)
y = dexp(x, rate = 0.2)
plot(x, y, type="l", lwd = 2, col = "red", ylab = "p")
x = seq(0, 4, length = 200)
y = dexp(x, rate = 0.2)
polygon(c(0, x, 4), c(0, y, 0), col = "lightgray")
```

The exponential distribution can be simulated in R with `rexp(n, lambda)` where `lambda` is the rate parameter. The mean or expected
value of the exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also also $\frac{1}{\lambda}$. 

## Theoretical Center

First, given values for the rate $\lambda$, we show where the distribution is centered at and compare it to the theoretical center 
of the distribution.  In this case, where $\lambda = 0.2$, the central tendency or mean is expected to be 
$E[x] = \frac{1}{\lambda} = \frac{1}{0.2} = 5$.

```{r pt1_1, echo = SHOW.ALL | TRUE, cache = FALSE}
lambda <- 0.2      # rate of descent
mu <- 1/lambda     # expected value 
sigma <- 1/lambda  # standard deviation 
```

Here, we illustrate through simulation that `rexp(n, lambda)` follows the law of large numbers, with cumulative means 
approaching $\frac{1}{\lambda}$ as the number of observations $n \rightarrow \infty$.

```{r fig1_1a, echo = SHOW.ALL | TRUE, cache = FALSE}
set.seed(12345)
nosim <- 1000

means <- cumsum(rexp(nosim, lambda)) / (1  : nosim)
g <- ggplot(data.frame(x = 1 : nosim, y = means), aes(x = x, y = y)) 
g <- g + geom_hline(yintercept = mu, size = 2, color = "blue") + geom_line(size = 1) 
g <- g + annotate("text", label = paste("E[x] = ", mu), x = 500, y = 1.15*mu, size = 8, colour = "blue")
g + labs(x = paste("Number of obs of rexp(), lambda = ", lambda, ", E[x] = ", round(mu, 3)), y = "Cumulative mean")
```

Now, we look specifically at the central tendency of the sample means from 1,000 trials of 40 observations each.

```{r fig1_1b, echo = SHOW.ALL | TRUE, cache = FALSE}
n <- 40 
x <- apply(matrix(rexp(n*nosim, lambda), nosim, n), 1, mean)
g <- ggplot(data.frame(x = x), aes(x = x)) + geom_histogram(alpha = .20, binwidth=.3, colour = "black", aes(y = ..density..)) 
g + geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed", size = 1.5) +
geom_text(aes(x = mean(x), label=paste("\n\nAverage of sample means = ", round(mean(x), 3)), y=0.25), 
color="red", angle=90, text=element_text(size=6)) +
geom_vline(aes(xintercept = mu), color = "blue", linetype = "dashed", size = 1.5) +
geom_text(aes(x = mu, label=paste("\n\n\n\nAverage of sample means = ", round(mu, 3)), y=0.25), 
color="blue", angle=90, text=element_text(size=6))
```

## Variability

Now, we show how variable the distribution is and we compare it to the theoretical variance of the distribution.  We know that
the population variance for the exponential function is given by $(\frac{1}{\lambda})^2$.  That gives us:

$$
    \sigma^2 = (\frac{1}{\lambda})^2 = (\frac{1}{0.2})^2 = 25
$$

```{r fig2_1a, echo = SHOW.ALL | TRUE, cache = FALSE}
n <- 40
x <- apply(matrix(rexp(n*nosim, lambda), nosim, n), 1, sd)
g <- ggplot(data.frame(x = x), aes(x = x^2)) + geom_histogram(alpha = 0.20, binwidth = 0.3, colour = "black", aes(y = ..density..)) 
g + geom_vline(aes(xintercept = mean(x^2)), color = "red", linetype = "dashed", size = 1.5) +
geom_text(aes(x = mean(x^2), label=paste("\n\n\nSample variance = ", round(mean(x^2), 3)), y = 0.05), 
color="red", angle=90, text=element_text(size=6)) +
geom_vline(aes(xintercept = sigma^2), color = "blue", linetype = "dashed", size = 1.5) +
geom_text(aes(x = sigma^2, label=paste("\n\n\n\nDistribution variance = ", round(sigma^2, 3)), y = 0.05), 
color="blue", angle=90, text=element_text(size=6))
```

## Normality

Through an experiment, we illustrate the approximate normality of the distribution by simulating of multiples of **`r nosim` trials** of 
sample means of `rexp(n, lambda)`, where $\lambda = 0.2$, where each trial contains **`r n` observations** each. 

```{r fig3_1, echo = SHOW.ALL | TRUE, cache = FALSE}
n <- 40 
cfunc <- function(x, n) sqrt(n) * (mean(x) - mu) / sigma
dat <- data.frame(
  x = c(apply(matrix(sample(rexp(n, lambda), nosim * 10, replace = TRUE), 
                     nosim), 1, cfunc, 10),
        apply(matrix(sample(rexp(n, lambda), nosim * 20, replace = TRUE), 
                     nosim), 1, cfunc, 20),
        apply(matrix(sample(rexp(n, lambda), nosim * 30, replace = TRUE), 
                     nosim), 1, cfunc, 30)
        ),
  size = factor(rep(c(10, 20, 30), rep(nosim, 3))))
g <- ggplot(dat, aes(x = x, fill = size)) + geom_histogram(alpha = .20, binwidth=.3, colour = "black", aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 2)
g + facet_grid(. ~ size)
```

## Coverage

Finally, we valuate the coverage of the confidence interval for $\frac{1}{\lambda}$: $\overline{X}\pm{1.96}\frac{s}{\sqrt{n}}$.  We 
know that a standard deviation of 1.96 corresponds to the 97.5 percentile.  We look at 
variations in $\lambda = {0.01, 0.02, ..., 0.25}$.

```{r fig4_1, echo = SHOW.ALL | TRUE, cache = FALSE}
n <- 40
lvals <- seq(0.01, 0.25, by = 0.01)
coverage <- sapply(lvals, function(l){
  phats <- rexp(n, l) / n
  ll <- phats - qexp(0.975) * sd(phats) / sqrt(n)
  ul <- phats + qexp(0.975) * sd(phats) / sqrt(n)
  mean(ll < l & ul > l)
})

ggplot(data.frame(lvals, coverage), aes(x = lvals, y = coverage)) + 
geom_line(size = 2) + geom_hline(yintercept = 0.975)
```

TBD.

### Evaluation notes

*Regarding the distribution of the mean of 40 exponentials. Did the student show where the distribution is centered at and compare it  to the theoretical center of the distribution?*

0: No<br />
1: Yes

*Regarding the distribution of the mean of 40 exponentials. Did the student show how variable it is and compare it to the theoretical variance of the distribution?*

0: No<br />
1: Yes

*Regarding the distribution of the mean of 40 exponentials. Did the student show that the distribution is approximately normal?*

0: No<br />
1: Yes

*Did the students evaluate  the coverage of the confidence interval for 1/lambda: $\overline{X}\pm{1.96}\frac{s}{\sqrt{n}}$*

0: No<br />
1: Yes

*Was the report created in knitr?*

0: No<br />
1: Yes

*Here's your opportunity to give this project +1 for effort. Did the student basically try to answer the question?*

0: No, it was really a cursory effort<br />
1: Yes, the student put in real effort and made an attempt t (regardless of correctness)

```{r finalize, echo = SHOW.ALL | FALSE}
# save working image so we can look at it later
save.image("~/Google Drive/projects/r/coursera/statinference-006/statinference-006-pt1.RData")
```



