---
title: "Statistical Inference Course Project, Part I"
author: "Allen Hammock"
date: "October 12, 2014"
output: html_document
---
```{r compile_opts, echo = FALSE}
    SHOW.ALL <- FALSE
    suppressMessages(library(knitr))
    opts_knit$set(verbose = SHOW.ALL, progress = TRUE)
```

```{r initialize, echo = SHOW.ALL | FALSE}
# set the working directory
setwd("~/Google Drive/projects/r/coursera/statinference-006")

# pre-load all required dependencies
load.package <- function(x) {
    if (!suppressMessages(require(x, character.only = TRUE))) {
        warning(paste("WARNING. Package", x, "not found. Installing now."))
        install.packages(x, dep = TRUE)
        if(!require(x, character.only = TRUE)) {
            stop(paste("ABORT. Package", x, "not found"))
        }
    }
}

required.packages <- c("dplyr", "ggplot2")
for(i in 1:length(required.packages)) {
    load.package(required.packages[i])   
}
```
# Background

## Part 1

### Problems
The exponential distribution can be simulated in R with `rexp(n, lambda)` where `lambda` is the rate parameter. The mean of exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also also $\frac{1}{\lambda}$. Set $\lambda={0.2}$ for all of the simulations. In this simulation, you will investigate the distribution of averages of 40 exponential(0.2)s. Note that you will need to do a thousand or so simulated averages of 40 exponentials.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponential(0.2)s. 

1. Show where the distribution is centered at and compare it to the theoretical center of the distribution.

```{r pt1_1, echo = SHOW.ALL | TRUE, cache = FALSE}
# solve a problem here
```

2. Show how variable it is and compare it to the theoretical variance of the distribution.

```{r pt1_2, echo = SHOW.ALL | TRUE, cache = FALSE}
# solve a problem here
```

3. Show that the distribution is approximately normal.

```{r pt1_3, echo = SHOW.ALL | TRUE, cache = FALSE}
# solve a problem here
```


4. Evaluate the coverage of the confidence interval for $\frac{1}{\lambda}$: $\overline{X}\pm{1.96}\frac{s}{\sqrt{n}}$.

```{r pt1_4, echo = SHOW.ALL | TRUE, cache = FALSE}
# solve a problem here
```

### Evaluation notes

*Regarding the distribution of the mean of 40 exponentials. Did the student show where the distribution is centered at and compare it  to the theoretical center of the distribution?*

0: No<br />
1: Yes

*Regarding the distribution of the mean of 40 exponentials. Did the student show how variable it is and compare it to the theoretical variance of the distribution?*

0: No<br />
1: Yes

*Regarding the distribution of the mean of 40 exponentials. Did the student show that the distribution is approximately normal?*

0: No<br />
1: Yes

*Did the students evaluate  the coverage of the confidence interval for 1/lambda: $\overline{X}\pm{1.96}\frac{s}{\sqrt{n}}$*

0: No<br />
1: Yes

*Was the report created in knitr?*

0: No<br />
1: Yes

*Here's your opportunity to give this project +1 for effort. Did the student basically try to answer the question?*

0: No, it was really a cursory effort<br />
1: Yes, the student put in real effort and made an attempt t (regardless of correctness)

```{r finalize, echo = SHOW.ALL | FALSE}
# save working image so we can look at it later
save.image("~/Google Drive/projects/r/coursera/statinference-006/statinference-006-pt1.RData")
```